---
title: "Assignment 2: Coding Basics"
author: "Andi Mujollari"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics, Part 1

1.  Generate a sequence of numbers from one to 30, increasing by threes. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r basics 1}
# My sequence will be from 1 to 30, added by 3
my_sequence <- seq(1,30, 3)
my_sequence

#With this code we tell R to find the mean value of the sequence we just created
mean_value <- mean(my_sequence) 
#With this code we tell R to find the median value of our sequence
median_value <- median(my_sequence)
#Here we ask R if mean is grater than the median
mean_grater_than_median <- mean_value > median_value
cat("Mean:", mean_value, "\n")
cat("Median:",median_value, "\n")
cat("Is mean grater than the median?", mean_grater_than_median, "\n")
#1. 

#2. 

#3. 

```

## Basics, Part 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r basics 2}
# In the first Vector i have assigned the names of the students
student_names <- c("Shila","Migel","Jona", "Uarda")
#In the second Vector i have assigned the test scores of each of them
student_scores <- c(98, 67, 76, 34)
#In the third Vector i see if they passed the test (TRUE or FALSE) with a passing grade of 50
passed_status <- student_scores >=50 
# In the end i collect the vectors into a data frame 
student_table <- data.frame(Name = student_names, Score = student_scores, Passed = passed_status)
student_table
```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: Data frames allows me to use different data types combined together. On the other hand, if i had to use a matrix instead, i would have to convert all the data in one type.

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement.

11. Apply your function to the vector with test scores that you created in number 5.

```{r basics 3}
#Here i will check if the students passed or failed the test by using if & else 
check_passing_score <- function(score) { 
  if (score >= 50) {
return(TRUE) 
    } else {
return(FALSE) }
}

for (score in student_scores) {
cat("Score:", score, "Passing:", check_passing_score(score), "\n")
}
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?
```{r}
#To compare the options first i have to try also the ifelse command 
check_passing_score <- function(score) {
  result <- ifelse(score >=50, "TRUE", "FALSE")
  cat(result, "\n")
}

 for (score in student_scores) {
cat("Score:", score, "Passing:", check_passing_score(score), "\n")
} 
```

> Answer: Both of the options worked!
